# GIS data models and file formats

## Data models

GIS data typically come in two data model types **vector** or **raster**.

### Vector data

The three basic vector data types are **points**, **lines** (also sometimes referred to as **polylines** or **linestrings**) and **polygons**. While they are treated as different data types, you can also consider them to be a nested hierarchy. For example, to make a line you need two or more points, while a polygon requires three or more lines. 

```{r vectors, fig.cap='The hierarchical construction of vector data types.', fig.asp=0.45, fig.align='center', echo=FALSE, warning=F, message=F}
library(tidyverse)
library(hrbrthemes)

data <- data.frame(set = factor(x=c("point", "line", "line", "polygon", "polygon", "polygon", "polygon"), levels = c("point", "line", "polygon")), x = c(1,1,2,1,2,1.5,1), y = c(1,1,1,1,1,2,1))

data %>%
 # tail(10) %>%
  ggplot(aes(x=x, y=y)) +
    geom_polygon(fill = "grey") +
    geom_path(color="black") +
    geom_point(shape=21, color="black", fill="#69b3a2", size=6) +
    theme_ipsum() + ylim(0.5,2.5) + xlim(0.5,2.5) +
    facet_wrap(.~set)

```

<br>

From this we can observe the different properties of the data types:

* a **point** is a _location_ in space defined by a set of coordinates based on a coordinate reference system (more about these later)
* a **line** is two or more points with straight lines connecting them, where each line has a _length_
* a **polygon** is a set of points connected by lines that form a closed shape, which has an _area_

Note that these "data types" are also commonly called _geometric primitives_ or _geometries_.

<br>

## File formats

As I mentioned before, there is a huge variety of file formats, which have proliferated as different software packages have developed their own set of "native" formats. Many of these are not open source.

You can view the list of most of the file types supported by R (or at least the types supported by library(rgdal)) by running the code `rgdal::ogrDrivers()`, which gives this output:

```{r, echo = F}
library(knitr) # To make the table with the kable() function
library(rgdal) # Holds the ogrDrivers() function

kable(head(ogrDrivers(), n = 80))
```

Note that there has been a big push to develop a standardized set of open source, efficient and interoperable file formats. 
