--- 
title: "A Minimal Introduction to GIS in R"
author: "Jasper Slingsby"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: rstudio/bookdown-demo
description: "This is a minimal introduction to GIS and handling spatial data in R compiled for the Biological Sciences BSc(Honours) class at the University of Cape Town."
---

# Overview

The goal is to give you a very brief introduction to Geographic Information Systems (GIS) in general and some familiarity with handling spatial data in R. GIS is a field of research that many people dedicate their entire lives to, yet we only have a week, so this really is a minimalist introduction. I'll focus on giving you a broad overview and some idea of how to teach yourself (using R).

The 5-day module includes daily ~1 hour lectures followed by self-study time where you will have daily goals/tasks to help you incrementally develop your own project. This project will be the deliverable for the module. I will be available for set office hours every afternoon to assist with these projects.

<br>

This is a _sample_ book written in **Markdown**. You can use anything that Pandoc's Markdown supports, e.g., a math equation $a^2 + b^2 = c^2$.

The **bookdown** package can be installed from CRAN or Github:

```{r eval=FALSE}
install.packages("bookdown")
# or the development version
# devtools::install_github("rstudio/bookdown")
```

Remember each Rmd file contains one and only one chapter, and a chapter is defined by the first-level heading `#`.

To compile this example to PDF, you need XeLaTeX. You are recommended to install TinyTeX (which includes XeLaTeX): <https://yihui.name/tinytex/>.

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
